### 2015-11-25 paper/1-introduction.tex
Tweak pqueue image, rewrite/reword a few subsections.

\subsection{Parallel computing}
%TODO parallel computing is necessary with more and more complex systems, blablabla (usual explanation)
Parallel execution of a DEVS simulation can lower runtime and increase the bound on the state space, thereby enabling simulation of more complex systems in the same time-frame.
While the shared memory parallelism offered by most modern hardware does not lower the state space bounds, it can reduce the runtime and offers more direct communication and control between entities involved in synchronization compared to distributed simulation. \\

@@

\subsection{Solution} %for lack of a better word
%TODO give our solution, motivating the need for a different kernel, instead of modifying another kernel (like adevs or PythonPDEVS)
The usage of the DirectConnect\cite{SymbolicFlattening} algorithm makes reusing the adevs kernel hard.
The dxexmachina kernel follows the design of the PythonPDEVS kernel where appropriate, but by the very nature of the implementation languages has to differ in key aspects (e.g. memory allocation strategy). 
The core aims of the project are to offer a deterministic simulation kernel where the simulation author is shielded as much as possible from the kernel implementation, without sacrificing performance. As in PyPDEVS, a model need be written only once for use in the different simulation kernels (with the exception of a non-trivial lookahead()).\\
The tracing framework from PyPDEVS was ported to allow optional verification of simulations. 
\subsection{Time}
The DEVS formalism has $\mathbb{R}$ as time base, but any implementation has to decide on a (finite) representation of time. Dxexmachina kernels can operate on IEEE754 floating point time units, or integral time. The latter significantly reduces the possible range of the simulation, but avoids possible approximation errors in floating point. Furthermore, the notion of $\epsilon$ as the absolute minimum between time points needs to be established, this is non-trivial for floating point. \\
The select() function, which arbitrates between concurrent events, is implicit in this project by providing any time representation with a secondary causality field.

% incomplete

>>>>

\subsection{Parallel computing}
%TODO parallel computing is necessary with more and more complex systems, blablabla (usual explanation)
Parallel execution of a DEVS simulation can lower runtime and increase the bound on the state space, thereby enabling simulation of more complex systems in the same time-frame.
While the shared memory parallelism offered by most modern hardware does not increase the state space bounds, it can reduce the runtime and offers more direct communication and control between entities involved in synchronization compared to distributed simulation. \\

@@

\subsection{Solution} %for lack of a better word
%TODO give our solution, motivating the need for a different kernel, instead of modifying another kernel (like adevs or PythonPDEVS)
The usage of the DirectConnect\cite{SymbolicFlattening} algorithm makes reusing the adevs kernel hard.
The dxexmachina kernel follows the design of the PythonPDEVS \cite{PythonPDEVS} kernel where appropriate, but by the very nature of the implementation languages has to differ in key aspects (e.g. memory allocation strategy). 
The core aims of the project are to offer a deterministic simulation kernel where the simulation author is shielded as much as possible from the kernel implementation, without sacrificing performance. As in PyPDEVS, a model need be written only once for use in the different simulation kernels (with the exception of a non-trivial lookahead).\\
The tracing framework from PyPDEVS was ported to allow optional verification of simulations. 
\subsection{Time}
The DEVS formalism has $\mathbb{R}$ as time base, but any implementation has to decide on a (finite) representation of time. Dxexmachina kernels can operate on IEEE754 floating point time units or integral time. In principle any type with well defined operators can be used as template parameter, but from a performance point of view a type fitting in a machine word offers obvious advantages.%ref to time paper.
An integral representation significantly reduces the possible range of the virtual time, but avoids approximation errors. Furthermore, the notion of $\epsilon$ as the absolute minimum between time points needs to be established, this is non-trivial for floating point.\\
The select() function, which imposes a sequential ordering between concurrent events, is implicit in this project by extending time representation with a secondary causality field. If A and B are imminent at time t then $t[1]_a < t[1]_b \oplus t[1]_b < t[1]_a$, while $t[0]_a == t[0]_b == t[0]$. This avoids evaluating the select() function on all imminent models, while still maintaining the deterministic order of concurrent transitions.

### 2015-11-25 paper/3-features.tex
Tweak pqueue image, rewrite/reword a few subsections.

%TODO explain that we are briefly going over the main features
\subsection{Based on PythonPDEVS}
%TODO say that it is based on PythonPDEVS
% link reference
The simulator is based on PythonPDEVS, and provides the following features: 
\begin{enumerate}
	\item Direct Connect
	\item Dynamic Structured DEVS
	\item Termination function. If specified, a termination function is applied every simulation round to each model to test whether the simulation can terminate. Only available in single-threaded simulation.
	\item State/Message can have any payload type. Different message types can be used together within the same simulation.
	\item Tracing An asynchronous, thread safe and versatile tracing mechanism allows exact verification of the simulation.
\end{enumerate}
The implementation tries to adhere to the C++ principle that you don't pay performance-wise for what you don't use. For this reason, the support for a termination function for the multi-threaded kernel was abandoned, as it is non-trivial to implement and had a non-negligible impact on the runtime, even when not in use.\\
The tracing is not comparable with adevs's listener interface. To be usable in optimistic simulation, the tracing of the simulation has to be reversible and only be committed at GVT points. Furthermore, the framework itself has to be threadsafe and deterministic so that a simulation will always produce the exact same output.
%TODO Decide whether we should list the functionality that is _not_ implemented
The following features from PythonPDEVS are not present
\begin{enumerate}
	\item Activity tracking and relocation
	\item Serialization
	\item Interaction 
	\item Distributed simulation
	\item Interactive control
\end{enumerate}
Serialization in this context is the ability to save/load a complete simulation to disk, not the state saving mechanisms required for TimeWarp.\\ State saving has no impact in a single threaded or conservative kernel.\\
Model allocation is done by a derivable allocator object which the user can implement to arrange a more ideal (domain-specific) allocation. If this is omitted, a default (non-activity-aware) allocation stripes the models over the simulation kernels.\\
Debugging tools such as a logger and a graph vizualizer are included which can track activity with respect to allocation for later study, but not online/dynamic as is possible in PyPDEVS.\\
While PythonPDEVS can be controlled from within a Python script and adevs has a Java interface, our implementation does not have any bindings to other languages.

@@

\subsection{Performance Improvements}
%TODO briefly discuss some of the additional changes you implemented
%TODO such as memory pools, memory allocators, profiling-guided-optimizations, ...
Continuous profiling of the kernels in several benchmarks highlighted the following key bottlenecks. 
% Maybe put more centrally that we trade memory for speed
\begin{enumerate}
	\item Heap allocation
	\item Strings 
	\item Scheduling
	\item Locking
	\item Smart pointers
\end{enumerate}
%integrate this in the list?
% In any case, this text is what we did, not what we should write so has to be condensed to the interesting parts (who except us likes bit shifting?)
% Pick what we want from this text and rewrite it
\subsubsection{Heap}
A kernel never sends a complete object to another kernel, only a pointer to the object.\\
The cost of heap allocations was minimized using thread\_local memory pools, combined with the (optional) usage of tcmalloc\cite{tcmalloc}.
This changed the ownership semantics of several objects in a non-trivial way, since the thread that creates an object has to destroy it (iff it can prove it is no longer used). Experiments with synchronized pools proved slower than malloc/free itself. \\
Replacing strings (in c++ : heap allocated variable sized arrays) with integer identifiers substantially decreased the runtime. This is a tradeoff, the user has more affinity with names than numbers as identifiers, so this translation is done mostly internal. A Model can be named using a std::string, but once the model is handed off to the kernel only integer identifiers remain (the same goes for ports, kernels, ...).
\\
\subsubsection{Avoiding Smart pointers}
While an important feature in C++11 in general, our initial usage of smart pointers for some types of objects was misplaced. Used across threads the reference counting becomes prohibitively expensive. Models will still be held by a smart pointer, as will a kernel, but a message is a raw pointer to compacted memory. Ideally a message object can (depending on payload) fit on a single cacheline.\\
\subsubsection{Locking}
Locking between kernels uses mostly atomic operations, where we can occasionally leverage memory orderings to only pay for synchronization when we need it. Messages are exchanged via a shared set of queues each with a dedicated lock.\\
On a higher level, we avoid the sending of synchronization messages entirely by writing the content (timestamp) directly into shared memory. This alleviates in part the potential delay/latency caused by null/eot messages being mixed with simulation messages in conservative implementations. 
\subsubsection{Schedulers}
PythonPDEVS has a wide range of scheduler for the user to choose from, with performance of each depending on the simulation type. Profiling showed in our case that, for a c++ implementation, the heap implementation used in adevs was faster than any of the schedulers we had tested before. Unlike most node based heaps, this scheduler uses a fixed size array where a heap is rebuilt or modified. Items are only updated, never removed.

>>>>

%TODO explain that we are briefly going over the main features
\subsection{Based on PythonPDEVS}
%TODO say that it is based on PythonPDEVS
The simulator is based on PythonPDEVS \cite{PythonPDEVS}  , and provides the following features: 
\begin{enumerate}
	\item Direct Connect \cite{SymbolicFlattening}
	\item Dynamic Structured DEVS
	\item Termination function. If specified, a termination function is applied every simulation round to each model to test whether the simulation can terminate. Only available in single-threaded simulation.
	\item State/Message can have any payload type. Different message types can be used together within the same simulation.
	\item Tracing An asynchronous, thread safe and versatile tracing mechanism allows exact verification of the simulation.
\end{enumerate}
The implementation tries to adhere to the C++ principle that you don't pay performance-wise for what you don't use. For this reason, the support for a termination function for the multi-threaded kernel was abandoned, as it is non-trivial to implement and had a non-negligible impact on the runtime, even when not in use.\\
The tracing is not comparable with adevs's listener interface. To be usable in optimistic simulation, the tracing of the simulation has to be reversible and only be committed at GVT points. Furthermore, the framework itself has to be threadsafe and deterministic so that a simulation will always produce the exact same output.
%TODO Decide whether we should list the functionality that is _not_ implemented
The following features from PythonPDEVS are not present
\begin{enumerate}
	\item Activity tracking and relocation
	\item Serialization
	\item Interaction 
	\item Distributed simulation
	\item Interactive control
\end{enumerate}
Serialization in this context is the ability to save/load a complete simulation to disk, not the state saving mechanisms required for TimeWarp.\\ State saving has no impact in a single threaded or conservative kernel.\\
Model allocation is done by a derivable allocator object which the user can implement to arrange a more ideal (domain-specific) allocation. If this is omitted, a default (non-activity-aware) allocation stripes the models over the simulation kernels.\\
Debugging tools such as a logger and a graph vizualizer are included which can track activity with respect to allocation for later study, but not online/dynamic as is possible in PyPDEVS.\\
While PythonPDEVS can be controlled from within a Python script and adevs has a Java interface, our implementation does not have any bindings to other languages.

@@

\subsection{Performance Improvements}
%TODO briefly discuss some of the additional changes you implemented
%TODO such as memory pools, memory allocators, profiling-guided-optimizations, ...
Continuous profiling of the kernels in several benchmarks highlighted the following key bottlenecks. 
\subsubsection{Heap}
A kernel never sends a complete object to another kernel, only a pointer to the object. This avoids a possibly expensive copy of the payload, but at the cost of allocation overhead.\\
This cost becomes prohibitively expensive in highly connected models, so to reduce that overhead we use thread\_local memory pools for states and messages, and optionally replace the system malloc with calls to tcmalloc\cite{tcmalloc}. In this way allocating threads do not block each other, and in a single threaded kernel we can leverage arena-style pools if desired.\\
This changed the ownership semantics of several objects in a non-trivial way, since the thread that creates an object has to destroy it (iff it can prove it is no longer used). Experiments with synchronized pools proved slower than malloc/free. \\
Initially the kernels used strings as identifiers, as is done in PyPDEVS, profiling quickly indicated this to be a performance bottleneck. C++ strings are heap allocated variable sized objects with a reference count. Access of that reference count across threads is expensive. The string implementation also calls malloc/free which are quite expensive. \\
String are more intuitive to work with from a user's standpoint, so as a compromise the user can reference models/ports by string name (usually when constructing the model). Once simulation starts all objects use integral identifiers for performance.
\\
\subsubsection{Smart pointers}
While an important C++11 feature in general, our initial usage of smart pointers for some types of objects was misplaced. Used across threads the reference counting becomes prohibitively expensive, and the (de)allocating caused significant contention between threads. Models are still held by a smart pointer, as is a kernel, but a message is a raw pointer to compacted memory. 
\subsubsection{Locking}
Locking between kernels uses mostly atomic operations, where we can occasionally leverage memory orderings to only pay for synchronization when we need it. Messages are exchanged via a shared set of queues each with a dedicated lock.\\
On a higher level, we avoid the sending of synchronization messages entirely by writing the content (timestamp) directly into shared memory. This alleviates in part the potential delay/latency caused by interleaving simulation with null/eot messages in conservative implementations. Optimistic still sends anti-messages as pointers, since the complete header of the message is required at the receiver side to annihilate the original. 
\subsubsection{Schedulers}
PythonPDEVS has a wide range of scheduler for the user to choose from, with performance of each depending on the simulation type. Profiling showed in our case that, for a c++ implementation, the heap implementation used in adevs was faster than any of the schedulers we had tested before. Unlike most node based heaps, this scheduler uses a fixed size array where a heap is rebuilt or modified in place depening on the nr of items to update. Items are only updated, never removed.
